---
title: "p8105_hw3_wl2926"
author: "Wenwen Li"
date: 2023-10-13
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(janitor)
library(ggridges)
# install.packages("devtools")
devtools::install_github("p8105/p8105.datasets")
library(p8105.datasets)
```

### Problem 1

#### Read in the data

``` {r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

#### Answer questions about the data

This dataset contains 1384617 rows and 15 columns, with each row
resprenting a single product from an instacart order. Variables include
identifiers for user, order, and product; the order in which each
product was added to the cart. There are several order-level variables,
describing the day and time of the order, and number of days since prior
order. Then there are several item-specific variables, describing the
product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs,
produce), and aisle (e.g. yogurt, fresh fruits), and whether the item
has been ordered by this user in the past. In total, there are 39123
products found in 131209 orders from 131209 distinct users.

Below is a table summarizing the number of items ordered from aisle. In
total, there are 134 aisles, with fresh vegetables and fresh fruits
holding the most items ordered by far.

``` {r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```


Next is a plot that shows the number of items ordered in each aisle.
Here, aisles are ordered by ascending number of items.

``` {r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


Our next table shows the three most popular items in aisles
`baking ingredients`, `dog food care`, and `packaged vegetables fruits`,
and includes the number of times each item is ordered in your table.

``` {r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady
Apples and Coffee Ice Cream are ordered on each day of the week. This
table has been formatted in an untidy manner for human readers. Pink
Lady Apples are generally purchased slightly earlier in the day than
Coffee Ice Cream, with the exception of day 5.

``` {r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


#Problem 2
```{r}
### Load libraries and data. 
library(p8105.datasets)
data("brfss_smart2010")
#print variable names
print(colnames(brfss_smart2010))
```

### Data Cleaning.    

```{r}
#Data cleaning
#clean variable names using janitor
brfss_smart2010 <- brfss_smart2010 %>%
  clean_names()
```

```{r}
#print unique levels of the "overall_health" variable
unique(brfss_smart2010$response)
unique(brfss_smart2010$topic)
```
### Filter data for the "Overall Health" topic and "Poor" to "Excellent" responses.   

```{r}
response_range <- c("Poor","Fair","Good","Very good","Excellent")
#filter for "Overall Health" topic and responses from "Excellent" to "Poor"
overallhealth_data <- brfss_smart2010 %>%
  filter(topic == "Overall Health" & response %in% response_range)

#organize responses as a factor with levels ordered from "Poor" to "Excellent"
overallhealth_data$response <- factor(overallhealth_data$response, 
                                      levels = response_range)
```

### Question 1: States observed at 7 or more locations in 2002 and 2010    
```{r}
states_2002 <- overallhealth_data %>%
  filter(year == 2002) %>%
  group_by(locationabbr, locationdesc) %>%
  summarise(num_locations = n()) %>%
  filter(num_locations >= 7) 
states_2002

states_2010 <- overallhealth_data %>%
  filter(year == 2010) %>%
  group_by(locationabbr, locationdesc) %>%
  summarise(num_locations = n()) %>%
  filter(num_locations >= 7)
states_2010
```
In the year 2002, none of the states were observed in 7 or more locations. The same case applies to the year 2010.  
### Question 2: Subset a dataset for "Excellent" responses and make a spaghetti plot.   

```{r}
excellent_data <- overallhealth_data %>%
  filter(response == "Excellent") %>%
  group_by(year, locationabbr) %>%
  summarise(avg_data_value = mean(data_value))

#spaghetti plot
ggplot(excellent_data, aes(x = year, y = avg_data_value, 
                           group = locationabbr, 
                           color = locationabbr)) +
  geom_line(size=1.5) +
  labs(title = "Mean data value for Excellent\n responses over time by State",
       x = "Year",
       y = "Mean data value",
       color="State") +
  theme_minimal()

```
      
There appears to be a generally decreasing trend over time for the average data value for "Excellent" responses in the "Overall health" topic.

There appears to be a generally decreasing trend over time for the average data value for "Excellent" responses in the "Overall health" topic.

### Question 3: Two-panel plot showing the distribution of "data_value" for responses ("Poor" to "Excellent") among locations in NY State for 2006 and 2010

```{r}
ny_data <- overallhealth_data %>%
  filter(locationabbr == "NY" & year %in% c(2006, 2010))

#two-panel plot
ggplot(ny_data, aes(x = data_value, fill = response)) +
  geom_histogram(binwidth = 2, position = "identity", alpha = 0.7) +
  facet_wrap(~year, ncol = 1) +
  labs(title = "Distribution of data value for responses (Poor to Excellent) in NY",
       x = "Data value",
       y = "Count",
       fill = "Response") +
  theme_minimal()
```
     
For NY, the year 2010 had a larger count of observations compared to the year 2006 in this subset. This included for observations under both the "Poor" and "Fair" responses on the overall health topic.  
For the year 2010, data values for the various responses were more consistent compared to the year 2006.

## Problem 3.    

### Read the data   

```{r}
#load accelerometer data
nhanes_accel <- read.csv("nhanes_accel.csv")
#display the structure of the dataset
dim(nhanes_accel)
```
     
```{r}
#load demographic data and skip first 4 rows
nhanes_covar <- read.csv("nhanes_covar.csv", skip = 4)
nhanes_covar <- na.omit(nhanes_covar)

#display the first four rows
str(nhanes_covar,4)
```
      
      
```{r}
#merge the data sets by "SEQN" (participant identifier)
nhanes_data <- merge(nhanes_accel, nhanes_covar, by = "SEQN")
#exclude participants less than 21 years of age
nhanes_data <- nhanes_data %>%
  filter(age >= 21)
#view dimensions of merged data
dim(nhanes_data)
```

```{r}
#convert variables to appropriate data types and order factors
nhanes_data$sex <- factor(nhanes_data$sex, 
                               levels = c(1, 2), 
                               labels = c("male", 
                                          "female"))
nhanes_data$education <- factor(nhanes_data$education, 
                               levels = c(1, 2, 3), 
                               labels = c("Less than high school", 
                                          "High school equivalent", 
                                          "More than high school"))
```


### Question 1: Table for the number of men and women in each education category.    
```{r}
edu_sex_data <- nhanes_data %>%
  group_by(education, sex) %>%
  summarise(count = n()) 
edu_sex_data_wide <- edu_sex_data %>% #convert table to wide format
  spread(key = sex, value = count)
kable(edu_sex_data_wide, 
      caption = "Number of Men and Women in each education category")
```
      

Female participants were the leading observations for the *`Less than high school`*and *`More than high school`*, whereas male participants led in the *`High school equivalent`* education categories.From the table, it is evident that the male participants had the larger proportion of observations in the dataset due to its having the larger difference in distribution for the *`High school equivalent`* education category on comparison with the differences in the other categories. 
      
### Question 2: Create a visualization of the age distributions for men and women in each education category.    

```{r}
ggplot(nhanes_data, aes(x = education, y = age, fill = factor(sex))) +
  geom_boxplot() +
  labs(title = "Age distributions by education and sex",
       x = "Education",
       y = "Age",
       fill = "Sex") +
  theme_minimal()
```
       
The age distribution of the participants whose education category was `Less than high school` shows left-skewed distributions for both sexes, as indicated by the median line of their respective box plots, which implies larger numbers of older individuals compared to the younger ones. The males in this category also had a larger range, while the females had a slightly higher median value.    
For both sexes of the participants whose education category was `High school equivalent`, their age values indicate left-skewed distributions, implying larger numbers of older individuals compared to the younger ones. The median age value of the females was larger compared to that of male participants. Again, their  the male participants were comparatively younger to their female counterparts, even though they had the larger number of observations as well as a wider range of age values.    
Lastly, for the participants whose education category was `More than high school`, the median values indicate right-skewed distributions, implying larger numbers of younger individuals compared to that of the older ones.  The male median age value is larger than that of female participants as well as the range of values.

     
      
### Question 3: Aggregate across minutes to create a total activity variable for each participant. Plot total activities against age, comparing men to women and having separate panels for each education level   

```{r}
#aggregate by sum foreach participant
nhanes_data$total_activity <- rowSums(nhanes_data[, 2:1441], na.rm = TRUE)

#Plot total activities against age by gender and education
ggplot(nhanes_data, aes(x = age, y = total_activity,color = factor(sex))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~education) +
  labs(title = "Total activity vs. age by gender and education",
       x = "Age",
       y = "Total activity",
       color = "Gender")+
  theme_minimal()
```
     
This plot compares the total activities against age for men and women, with separate panels for each education level. For the `Less than high school` education category, the trend lines show younger females as more active compared to younger males. At about 45 years of age, their total activities were about equal, and males were more active thereafter. For the other education categories, females show more activeness compared to males. Across all the education and sex categories, the general trend is that the total activity declines over age. The rate of decline of the total activity is highest for the `Less than high school`, and least for the `More than high school`.  

### Question 4: Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex.    

```{r}
#subset columns 2 to 1441, which represent the minutes of the day
minutes_data <- nhanes_data[, 2:1441]

#create a new data frame with hourly activity aggregates
hourly_data <- data.frame(sex = nhanes_data$sex, 
                          education = nhanes_data$education)
#loop through the columns and aggregate every 60 columns=1 hour
hour_labels <- sprintf("hour_%d", 1:24)  #hour labels
for (i in seq(1, ncol(minutes_data), 60)) {
  #calculate the activity sum for each hour, add to the new data frame
  hourly_data[hour_labels[i %/% 60 + 1]] <- 
    rowSums(minutes_data[, i:(i + 59)], na.rm = TRUE)}
#print the first few rows of the new data frame
str(hourly_data)
```

```{r}
#reshape to long format
hourly_data_long <- hourly_data %>%
  pivot_longer(cols = starts_with("hour_"), 
               names_to = "hour", 
               values_to = "hourly_activity") %>%
  mutate(hour=as.numeric(str_extract(hour, "\\d+")))#extract integer part
#preview long format
head(hourly_data_long)
```


```{r}
#plot 24-hour activity time courses by education level and sex
ggplot(hourly_data_long, aes(x = as.numeric(hour), 
                             y = hourly_activity, color = factor(sex))) +
  geom_point(alpha = 0.5) + 
  scale_x_continuous(breaks = seq(0, 24, by = 4), 
                     labels = as.character(seq(0, 24, by = 4))) + 
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~education) +
  labs(title = "24-Hour activity time courses by education and sex",
       x = "Hour of the day",
       y = "Hourly total activity",
       color = "Sex") +
  theme_minimal()
```
     
This three-panel plot shows the 24-hour activity time courses for each education level, using color to indicate sex. For the `Less than high school` education category, the male and female hourly activity overlap at different times of the day, with peak activity occurring at around noon time. For the `High school equivalent` education category, there is a similar peak of activity at around noon time, but the female participants were consistently more active after around the 4th hour of the day. Lastly, for the `More than high school` education category, the female participants were also consistently more active after around the 3rd hour of the day. Within this category were also the largest number of outliers in terms of hourly activity for both sexes. Notably, after the 12th hour of the day, the loess curve of the hourly activity for the `More than high school` is more flat compared to the other education categories, indicating more activeness.


